Monte Carlo solvers for sparse linear systems

### 2015-09-16 MAX_reports/MC_solvers_article/MC_solvers.tex
Breaking paper into multiple files for easier editing/collaborating.

\section{Introduction}
For a long period the
pursuit of a reliable solver for linear and nonlinear algebraic systems has
been one of the most important issues in Numerical Analaysis. The
mathematical properties of the model (e.g. likely sparsity pattern of the
matrix) have urged scientists to look for efficient and pattern-driven
solvers, capable to cope with curse of dimensionality and high scalability
requests.

As scalability computing move towards exascale facilities (which stands for
machines computing up to $10^{18}$ flops), new numerical schemes suitable for
this
kind of architectures are strongly demanded. In fact the purpose is to combine
high
fidelity of the results with an optimized use of hardware resources at hand.

Recently new algorithms that combine statistical and numerical techniques have
been developed. Even though \textit{numerical efficiency} has been sacrificed
sometimes, the advantages in terms of \textit{robustness} of these methods make
it worth carrying on studies in this direction.
A class of methods that may be employed for this purpose is represented by
Monte Carlo linear solvers.

These algorithms are the main topic covered in the work presented here,
starting from what was developed so far by some of
the authors who gave significant contributions (see \cite{Hal1962},
\cite{Hal1994},
\cite{DA1998}, \cite{DVA2001}, \cite{AADBTW2005},\cite{ESW2013} and
\cite{EMSH2014}). \newline

As underlined in \cite{DA1998}, Monte Carlo methods may be split into two
classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and
\textit{iterative methods} (\cite{Hal1962},
\cite{Hal1994}, \cite{ESW2013}
and \cite{EMSH2014}). The first are characterized by a merely stochastic
scheme,
therefore the provided error with respect to the exact solution is made of
just a stochastic component. The second ones, instead, combine numerical and
statistical schemes generating two types of error: a
\textit{stochastic} one and a \textit{systematic} one. It does not
mean
that
it will be always simple to recognize them separately. However it is important
to
be
aware of this intrinsic structure, in order to target what is the part of the
scheme that requires a refinement (e.g. increase of numerical iterations
rather than random walks).\newline

The paper is organized as follows. In Section 2 we describe all the
viable stochastic settings for the definition of the Markovian process. In
Section 3 we reinterpret probabilistic facts in terms of spectral properties of
the linear system to be solved. Section 4 is dedicated to necessary conditions
and sufficient conditions for the convergence of the algorithms. In Section 5
different adaptive approaches are presented, in order to truncate the random
walks and to decide how many samplings to use. Section 6 regards the study
of matrices for which the convergence of the stochastic schemes is facilitated.
Then in Section 7 a hybrid scheme combining numerical and probabilistic
techniques is described: the \textit{Monte Carlo Synthetic Acceleration}.
Numerical results are presented in Section 8 and 9, while Section 10 is
dedicated to
conclusive remarks and goals for the future.

\section{Stochastic Linear Solvers}

We consider the solution of systems of linear equations of the form
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in
\mathbb{R}^n$.

By applying a left preconditioning, the system (\ref{linsys}) gets the
following shape:
\begin{equation}
P^{-1}A \mathbf{x}=P^{-1}\mathbf{b},
\label{linsys_prec}
\end{equation}
(\ref{linsys_prec}) can be
reinterpreted as a fixed point scheme
\begin{equation}
 \mathbf{x}=H\mathbf{x}+\mathbf{f}.
 \label{fixedpoint}
\end{equation}
where $H=I-P^{-1}A$ and $\mathbf{f}=P^{-1}\mathbf{b}$.

Assuming that the spectral radius $\rho(H)<1$, the solution to
(\ref{fixedpoint}) can be written in terms of a power series of
$H$ (Neumann series)
\[
\mathbf{x}=\sum_{i=0}^\infty H^i\mathbf{f}.
\]
Therefore the fixed point scheme generates a sequence of approximate solutions
$\{\mathbf{x}^{(k)}\}_{k=0}^{\infty}$ which converges to the exact solution
regardless of the initial guess $\mathbf{x}_0$.

By restricting the attention to a single component of $\mathbf{x}$ we
have
\begin{equation}
x_i=\sum_{\ell=0}^\infty \sum_{k_1}^n\sum_{k_2}^n\cdots \sum_{k_{\ell}=1}^n
H_{k_0,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator
defined on a random walk.

Let us start considering a random walk whose
state space $S$ is characterized by the set of indices of the forcing term
$\mathbf{f}$:
\[
S=\{1,2,\cdots, n\} \subset \mathbb{N}.
\]
Each $i$-th step of the random walk has a random variable
$k_i$ associated with it. The realization of $k_i$ represents the index of the
component of $\mathbf{f}$
which is visited in the current step of the random walk.

The way the transition probability is built and the way the initial state of
the random walk is chosen give birth to two different approaches we are
going to use: the \textit{Forward} and \textit{Adjoint} methods.

\subsection{Forward Method}

The goal is to evaluate a functional such as
\[
J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
We can use it to build the initial probability $\tilde{p}:
S\rightarrow [0,1]$ of the random walk such that
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert
h_i\rvert}.
\]
It is important to stress out that the role of vector $\mathbf{h}$ is
restricted to the construction of the initial probability. Once that is done,
it is
not used anymore in the definition of the stochastic process.
A possible choice for the transition probability $P$ can be
\[
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n
\lvert H_{i,k}\rvert}.
\]
where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
A related sequence of random variables $w_{i,j}$ can be defined
such that
\[
w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
\]
The probability distribution of the random variables $w_{i,j}$ is represented
by the transition matrix that rules the stochastic process. The $w_{i,j}$'s
just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_j$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad
j=1,\cdots, i.
\]
By defining
\[
X(\nu)=\sum_{m=0}^k W_m f_{i_m}
\]
as the random variable associated with a specific permutation $\nu$, we can
define the estimator $\theta_i$ such as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
\]
The integer $n$ represents the size of the solution vector to (\ref{linsys})
and
the
index
$i$
is referred to the component of the solution vector we want to compute.
$P_{\nu}$ is the probability associated with a specific permutation of the
random walk.
It can be proved that
\[
E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
\]
and
\[
\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
\]

A possible choice for $\mathbf{h}$ is a vector of the standard basis. This
would correspond in setting manually the initial state of the random walk,
which turns the related initial probability into a Kronecker delta
\[
\tilde{p}(k_0=i)=\delta_{i,j}.
\]
By doing so, we have
\begin{equation}
\theta_i=E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=x_i=\sum_{l=0}^\infty
\sum_{k_1=1}^{n}\sum_{k_2=1}^n\cdots \sum_{k_{\ell}=1}^n
P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},
k_{\ell}}w_{k_0,k_1}w_{k_1,k_2}\cdots
w_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{dir_mean}
\end{equation}

As regards the variance, we remember that the following relation holds:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg] - \bigg (E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]\bigg )^2.
\label{dir_var}
\end{equation}

Hence the variance can be computed as the difference between the second
moment of the random variable and the square of the first moment.\newline

In order to apply the Central Limit Theorem (CLT) to the estimators defined
above, we must require that
the estimators have both finite expected value and variance. This is
equivalent in
checking the finiteness of the expected value and of the second moment.
Therefore we have to impose the following conditions:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}\bigg]<\infty
\end{equation}

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_{\ell}}^2\bigg]<\infty
\end{equation}

The Forward method presented above, however, has the limitation of employing an
entire set of permutations to estimate just a single entry of
the solution at a time. Therefore, in order to estimate the entire solution
vector for \ref{linsys_prec}, we have to employ as many sets of permutation as
the number of entries composing the vector. Of course this might turn to be
extremely expensive both in terms of time and resources for most of the
challenging problems coming from applications. This is essentially the reason
why usually it is preferred to adopt another stochastic algorithm which is
explained here following: the \textit{Adjoint} Monte Carlo method.

\subsection{Adjoint Method}

A second Monte Carlo method can be derived by considering the linear system
adjoint to (\ref{linsys_prec})
\begin{equation}
(P^{-1}A)^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.

By reformulating the fixed point scheme, introducing initial probability,
transition probability and weights
sequences in the same fashion as done before, the expected value for the
estimator becomes
\begin{equation}
\theta_j=E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
j}\bigg]=\sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2}^n\cdots\sum_{k_{\ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}\delta_{k_{\ell},j}.
\label{adj_mean}
\end{equation}
This estimator is known in literature as \textit{collision} estimator.

We can notice now that the component of the source vector, in each
permutation of the random walk, is associated with the initial step. The
Kronecker delta
at the end of the series stands for a filter. It means that if the goal is to
estimate the $j$-th component of the solution vector, just a subset of all the
permutations gives a contribution to this. The only permutations considered are
those ones that currently reside on the index associated with the entry to be
estimated.

The variance is

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_0}\delta_{k_{\ell},j}\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2\delta_{k_{\ell},j}\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
f_{k_0}\delta_{k_{\ell},j}\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

On the same lead of what done for the Forward method, there is need to impose
finiteness of the expected value and second moment. Therefore the following
conditions must be verified:

\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}\delta_{k_{\ell},
j}\bigg]<\infty \quad j=1,\cdots,n
\end{equation}
and
\begin{equation}
 E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2\delta_{k_{\ell},j}\bigg]<\infty \quad j=1,\cdots,n.
\end{equation}

The main advantage of this method, compared to the Forward one, consists in the
fact that a single set of permutation is used to estimate the entire solution
vector. Therefore, in general, this algorithm has to be preferred to the
previous one since it enables to reduce dramatically the computational burden.

In literature another estimator is employed along with the Adjoint Monte Carlo
method, the so called \textit{expected value} estimator. Its
formulations is as follows:

\begin{equation}
\theta_j=E\bigg[f_j + \sum_{\ell=0}^\infty
W_{\ell}H_{k_{\ell}, j}^T\bigg]=f_j
+ \sum_{\ell=0}^{\infty}\sum_{k_1}^n\sum_{k_2} ^n\cdots\sum_ { k_ { \ell}}^n
f_{k_0}P_{k_0,k_1}P_{k_1,k_2}\cdots P_{k_{\ell-1},K_{\ell}}w_{k_0,k_1}\cdots
w_{k_{\ell-1},k_{\ell}}H_{k_{\ell},j}^T.
\label{adj_mean1}
\end{equation}

Differently from the collision estimator, the expected value estimator averages
the deterministic contribution of the iteration matrix over all the potential
states $j$ that might be reached from the current state $\ell$. The variance
in this case becomes:

\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
f_{k_0}H_{k_{\ell},j}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
f_{k_0}^2 {H_{k_{\ell},j}^T}^T\bigg ] - \bigg (E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
f_{k_0}H_{k_{\ell},j}^T\bigg]\bigg )^2\quad j=1,\cdots,n
\label{adj_var}.
\end{equation}

In the sections dedicated to the numerical tests, both the collision estimator
and the expected value are employed, in order to test their effectiveness.

\subsection{Hybrid Stochastic/Deterministic Methods}

Even if the standard Monte Carlo, in both its Forward and Adjoint formulation,
has the advantage of being embarrassingly parallelizable (because of the
independence of the samplings employed), it has a drawback as
well. In fact its slow rate of convergence, associated with the
$\frac{1}{\sqrt{N}}$ behavior predicted but the CLT, prevents its use for
problems of actual interest in applied linear algebra (with many d.o.f's).

Moreover, when the spectral radius of the iteration matrix is close to the
unit
value, Monte Carlo direct methods reach convergence after an excessive number
of steps for each
permutation. In particular this aspect will be pointed out more by some
numerical tests analyzed in the upcoming sections of this work.
This motivation makes the standard Mote Carlo paradigm even more
unpractical for solving linear systems associated with realistic applications.
\newline

To cope with this, in the early 2000's Evans et al. introduced a new method:
the
\textit{Monte Carlo Synthetic Acceleration (MCSA)}
(\cite{ESW2013} and \cite{EMSH2014}). Earlier contributions on these direction
are attributed also to other scientists (e.g. \textit{Sequential Method} by
J.H. Halton). However the MCSA is the algorithm we are going to focus on more
hereafter. If
we think about a fixed point formulation of the problem
such as $\mathbf{x}=H\mathbf{x}+\mathbf{b}$, then the Monte Carlo Synthetic
Acceleration assumes this form

\begin{algorithm}[H]
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$x_{num}$}
 $\mathbf{x}^{l}=\mathbf{x}_0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
  $\delta \mathbf{x}^{l+\frac{1}{2}}=(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$\;
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
\mathbf{x}^{l+\frac{1}{2}}$\;
 }
 $x_{num}=x^{l+1}$\;
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}
The Monte Carlo method is used to compute the updating contribution $\delta
\mathbf{x}^{l+\frac{1}{2}}$.

The role of the first line of the iterative process has the goal to compress
the residual of the previous iteration. This is the actual difference
between the MCSA and the Sequential Monte Carlo. Once this done, the corrected
residual
is employed to compute an update $\delta x^{1+\frac{1}{2}}$ to the solution via
one of the Monte Carlo techniques introduced before. The utility of compressing
the residual at the very beginning of the iteration will be shown in the two
last test cases for numerical validations.
The intent of MCSA is to designate most of the time employed by the
computation to the accomplishment of the stochastic part of the algorithm.
This
is useful to decrease the probability of fault occurrence in the non-resilient
steps, which is represented by the non-stochastic component of the numerical
scheme presented above. In particular, the higher is the percentage of time
spent on the computation of $\delta x^{1+\frac{1}{2}}$ via Monte Carlo, the
better it should be in terms of resilience.

\subsubsection{Sequential Monte Carlo}

\subsubsection{Monte Carlo Synthetic Acceleration}

\section{Convergence Behavior of Stochastic Methods}

The requests imposed on the expected value and second moment of the estimators
can be reformulated in a purely deterministic setting by looking at the
spectral radii of some properly defined matrices.

For instance the condition of finiteness of the expected value can be
reformulated by requiring
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}

where $H$ is the iteration matrix of the fixed point scheme.
In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered
as
power series in terms of $H$ and the spectral radius of $H$ smaller
than 1 is a necessary and sufficient condition for the Neumann series to
converge.

By analogy we aim at reproducing a similar reasoning for the second moment of
the estimator, since we want it to be finite as well.
By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
for
the Forward and the Adjoint method respectively, we notice that the second
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:

\[
 \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
\]

or

\[
 \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
\]

Accordingly to the necessary and sufficient condition for the power series to
converge, we require
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}

Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
the stochastic schemes presented in this work. Having two conditions to
verify makes the numerical treatment of the linear system via MC very
difficult. We will discuss this in details later in the treatise. \newline

\subsection{Necessary and Sufficient Conditions}

Here below some results presented in
\cite{MASC2013} about necessary conditions and sufficient
conditions for
convergence are enunciated. In particular a suitable choice for constructing a
transition probability is discussed to guarantee and speed-up convergence.

We remember that the construction of the transition probability must satisfy
some constraints such as

\[
\begin{cases}
  P_{i,j}\ge 0 \\
 \sum_{j=1}^N P_{i,j}=1
\end{cases}
\]

plus one more restriction which varies depending on the particular Monte Carlo
approach chosen

\begin{align*}
& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0
\end{align*}.

The previous restrictions on the attainable values for the entries of the
transition probability are called \textit{transition conditions}.

At first an introductory theoretical result is shown, necessary to motivate the
choices made afterwards for the choice of the transition probability.

\begin{thm}
 Consider a vector $a=(a_1,a_2,\cdots,a_N)^T$ where at least one element is
nonzero, $a_k\ne0$ for some $k\in\{1,\cdots,N\}$.
\begin{itemize}
 \item For a probability vector $p=(p_1,p_2,\cdots,p_N)^T$ satisfying the
transition conditions, the lower bound of $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is $\bigg(\sum_{k=1}^N \lvert a_k\rvert\bigg)^2$
\item There always exists a probability vector such that $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}\ge c\ge 1$ for all $c>1$.
\end{itemize}
\label{lemma}
\end{thm}

It might be observed that the minimum of the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as
much
as possible. \newline

For the sake of simplicity let us introduce now a generic random walk
truncated at a certain $k$-th step
\[
 \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow
r_k
\]
and the statistical estimator
\[
 X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
\]
which is associated with the Forward Method.

Then the following theorem holds
\begin{thm}\textit{(Suited to the Forward Method)}

Let $H\in \mathbb{R}^{n\times n}$ such that $\lVert H\rVert_{\infty}<1$.
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
$k$-th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{for_thm}
\end{thm}

If we introduce the estimator
\[
 X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
\]
which is associated with the Adjoint Method, then we can state a
theorem specular to \ref{for_thm}.

\begin{thm}\textit{(Suited to the Adjoint Method)}

 Let $H\in \mathbb{R}^{n\times n}$ with $\lVert H\rVert_{1}<1$.
Consider $\nu_k$ as the realization of a random walk $\gamma$ truncated at the
$k$-th step. Then,
there always exists a
transition matrix $P$ such that
$Var\Big(X(\nu_k)\Big)\rightarrow 0$ and
$Var\Big(\sum_{\nu}X(\nu_k)\Big)$ is bounded as $k\rightarrow \infty$.
\label{adj_thm}
\end{thm}

In particular, relying on what guaranteed by Theorem \ref{lemma}, in both
Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO
transition probability.

These theoretical results represent sufficient conditions for the convergence
of the
Forward and Adjoint Monte Carlo and they can be easily checked.
However requiring $\lVert
H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding.
Indeed, even if there exist some preconditioners able to guarantee this
condition (e.g. approximate inverse preconditioners with a proper value of
dropping
tolerance), it may entail to compromise the sparsity of the
iteration
matrix.

Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and
\ref{adj_thm} are too demanding to be verified, the necessary and sufficient
condition
on the
spectral radius of $\hat{H}$ is the only criterion to take as reference.
Nonetheless this constraint consists in
computing the spectral radii of two matrices and the calculation
might be very costly. Moreover the computation of the spectral radii is
affected by resilience issues as well as
a deterministic solver for the resolution of
a linear system. Therefore the original difficulty is transferred without being
actually tackled.

\subsubsection{Forward MC}

\subsubsection{Adjoint MC with Collision Estimator}

\subsubsection{Adjoint MC with Expected-Value Estimator}

\subsection{Construction of Transition Probabilities}

The way the transition probability is defined is extremely important. In fact
in some circumstances it might determine whether the stochastic scheme
converges or not. Two very simple ways of defining it are listed here
below. These are the approaches adopted typically in literature.

\subsubsection{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements
in each row have equal probability of occurring.
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing
probabilities to all the entries of the transition probability accordingly
to the magnitude of the elements. For instance we may employ the following
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{i,j}\rvert^p}{\sum_{k=1}^n
\lvert H_{i,k}\rvert^p}
\]

\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{j,i}\rvert^p}{\sum_{k=1}^n
\lvert H_{k,i}\rvert^p}
\]

where $p\in \mathbb{N}$.
This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005},
is called
\textit{Monte Carlo Almost Optimal} (MAO).
The reason why the almost optimal probability is called this way is associated
with a notice from Theorem \ref{lemma}. Indeed from
the mentioned lemma we know that the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. This implies that the almost optimal probability minimizes the
$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$
in the
Adjoint method. This is very important to be pointed out since these two norms
are so far the best practical upper bounds for $\rho(\hat{H})$ that we can
easily
handle. In the Section dedicated to the numerical results it will be proved
that the Almost Optimal Probability eases the convergence of the MC solver,
even in cases when the employment of the Uniform Probability does not provide
convergence at all.

\subsection{Classes of Matrices with Guaranteed Convergence}

As we have already seen, sufficient conditions for
the convergence of the Monte Carlo linear solvers are very restrictive (see
\cite{MASC2013}), whereas the necessary and sufficient condition requires the
computation of
$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a
proper
preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may
be always guaranteed, whilst $\rho(\hat{H})<1$ is more
problematic to be verified.

In order to bypass the cost of these checks, it is useful to pursue for sets of
matrices for
which the Monte Carlo linear solver always converges, maybe with the use of an
ad-hoc preconditioner.

\subsubsection{Strictly Diagonally Dominant}

 One of the these sets is
represented by strictly diagonally dominant matrices.

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
by rows if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{i=1\\ i\ne j}}^{i=n}\lvert
a_{ij}\rvert
\label{sddr}
 \end{equation}
\end{defn}

\begin{defn}
 A matrix $A\in\mathbb{R}^{n\times n}$is strictly diagonally dominant (s.d.d)
by columns if
 \begin{equation}
    \lvert a_{ij}\rvert>\sum_{\substack{j=1\\ j\ne i}}^{j=n}\lvert
a_{ij}\rvert
\label{sddc}
 \end{equation}
\end{defn}

When (\ref{sddr}) holds we can resort to a left diagonal preconditioning and we
get an iteration matrix $H=I-D^{-1}A$ such that $\lVert H \rVert_{\infty}<1$.

Introducing a MAO transition probability for the Forward Method we get
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
\]
therefore the entries of $\hat{H}$ are defined as follows
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
(\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ).
\]
Therefore
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ) =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
\]
This automatically implies that $\rho(\hat{H})<=\lVert
\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
However nothing a priori can be
stated about the behavior of the Adjoint method.\newline

Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning
and we
get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
In this case, by following a similar reasoning to the one made before, we
conclude that the Adjoint Method convergences thanks to $\lVert
\hat{H}\rVert_1<1$,
whereas nothing can be said in advance as concerns the Forward method.

\subsubsection{Generalized Diagonally Dominant}

 Another set of matrices for which the convergence of
the MC solvers is guaranteed is made of \textit{generalized diagonally
dominant matrices}.

\begin{defn}
A square matrix $A\in\mathbb{R}^{n\times n}$ is said to be generalized
diagonally dominant if
\[
 \lvert a_{ii}\rvert x_i \ge \sum_{\substack{j=1\\j\ne i}}^n \lvert
a_{ij}\rvert
x_j, \quad i=1,\cdots,n
\]
for some positive vector $\mathbf{x}=(x_1,\cdots,x_n)^T$.
\end{defn}

A proper subset of the generalized diagonally dominant matrices is represented
by $M$- matrices (see \cite{Ax1996}).
There are many definitions of $M$-matrices that may be found on books and all
of them are equivalent. The one presented in \cite{Saad} states the
following

\begin{defn}
A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it
satisfies the following four properties:
\begin{itemize}
 \item $a_{i,i}>0$ for $i=1,\cdots,n$
 \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
 \item $A$ is nonsingular
 \item $A^{-1}\ge 0$
\end{itemize}
\end{defn}

A very important result about $M$-matrices enables to come up with a convergent
splitting and, therefore, a convergent fixed point scheme.

In particular Jacobi, Block Jacobi and Gauss Seidel  are
convergent
splitting for an $M$-matrix.
However, the construction of a splitting like this is not enough to ensure the
convergence of
the Monte Carlo linear solver. In fact none of these conditions is sufficient
to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
be considered.\newline

For this purpose we
appeal to a result in \cite{Li2002}. In this paper the author presents an
algorithm to transform a generalized diagonally dominant matrix with
nonzero diagonal entries into strictly diagonally dominant. The
algorithm works for a generic complex matrix $A\in\mathbb{C}^{n\times n}$.

Here below we report the algorithm at hand\newline

\begin{algorithm}[H]
 For a given complex matrix $A$, $a_{ii}\ne 0$, $i=1,\cdots, n$;\newline
 \begin{enumerate}
  \item Compute $S_{i}=\sum_{\substack{j=1 \\ j\ne i}^{n}}\lvert
a_{ij}\rvert$,
$i=1,2,\cdots,n$
\item Set $t=0$. For $i=1,2,\cdots, n$, if $\lvert a_{ii}\rvert>S_i$, then set
$t=t+1$
\item If $t=0$, then print "A is not a GDDM": END
\item If $t=n$, then print "A is a GDDM": END
\item \For{i=1,n}{$d_{i}=\frac{S_i+\varepsilon}{\lvert
a_{ii}\rvert+\varepsilon} \quad
\varepsilon>0,\quad j=1,2,\cdots,n$\;
$a_{ji}=a_{ji}\cdot d_i$}
\item Go to step 1.
 \end{enumerate}
 \caption{Algorithm to turn a GDDM matrix into a s.d.d by rows.}
\end{algorithm}

This approach turns a generalized diagonally dominant matrix into a
strictly diagonally dominant matrix by rows. By substituting
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
at step 1 with
$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
and by replacing
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
that turns a GDDM matrix into a s.d.d. by columns.\newline

Once we have applied this transformation to the matrix at hand into a s.d.d.,
we can use the Monte Carlo linear solver which is ensured to converge.

\subsubsection{Block Diagonally Dominant}

 In this section we wonder in which situations a block diagonal preconditioning
might come up with a convergent Monte Carlo linear solver.
By mimicking the computations we already showed previously, the iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
preconditioning takes the form

>>>>

\input{intro}
\input{mcls}
\input{convergence}
\input{results}
\input{conclusion}
